Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-05-08 paperDistance/paper.tex
edited toy model section

In this paper we use the noise deconvolved observed \cmd\ as a prior to infer more precise parallax pdfs from \gaia\ measurements.
Understanding this method is challenging because we are using the data both for the prior and the likelihood in our inference.
To ease understanding, here we will demonstrate this method with a simpler toy model.
Instead of inferring more precise parallax pdfs using the 2D distribution of the \cmd\ as our prior, for this simpler model we will infer more precise pdfs of the value $Y$.
We will do this simpler inference using a prior that is also a 2D distribution: a 1D Gaussian in the y-direction, with a running mean $\mu = mx + b$ and some thickness $t$.
Here the thickness of the distribution mimics the true thickness of the \cmd\ due to age, mass and metallicity.
The $1\sigma$ contours as well as (noise free) samples from the toy model distribution are shown in Figure \ref{fig:toy}, upper left panel.
Similarly, noisy samples from this distribution are shown in the upper right panel.
Using a representation of the true distribution (which we learn from the noisy data) as our prior, we can infer some posterior belief of $y_{\true,n}$, a true sample of the distribution, from $y_n$, a noisy sample of the distribution.

Setting up the problem in more detail, we have the true underlying distribution of points and the noisy data

@@

Here is a good point to step back and acknowledge that our method is odd/strange/different/new (which maybe you're also currently thinking but can't quite express why yet). We are using the set of data to determine the noise-deconvolved distribution, and then using that noise-deconvolved distribution to make inferences on the data. We are technically using the data twice! If we were going to be fully, technically correct, we should learn the distribution, but leave one datum out; the one datum we want to infer the true y value for\footnote{DWH: Remember what this brain dump is for}. Leaving one datum out to estimate the prior and then inferring the posterior for that datum is way too expensive for the 1024 points in our toy model, and also for the 1.2 million stars in our actual data set. Luckily with large data sets, even at 1024, 1 datum is $\sim0.1\%$ of your dataset and therefore leaving that point out will not change the MML parameters of your noise-deconvolved distribution, and with the \gaia\ dataset of over a million stars, we can safely step into the Empirical Bayes regime. (Talk more about emperical bayes here?)

So using Bayes theorem
\begin{eqnarray}
p(y_{\true,n} \given y_n, \tilde{\sigma}_n) &=& p(y_n \given y_{\true,n}, \sigma_n)\,p(y_{\true,n}) \\
p(y_n \given y_{\true,n}, \sigma_n) &=& \mathcal{N}(y_n \given y_{\true,n}, \sigma_n) \\
p(y_{\true,n}) &=& \mathcal{N}(y_{\true,n} \given \hat{m}\, x_n + \hat{b}, \hat{t})
\label{eq:toyBayes}
\end{eqnarray}

Since everything is Gaussian, this can be done analytically by multiplying Gaussians together. In Figure \ref{fig:toy} lower left panel, the blue points represent the posterior belief of the true value $p(y_{\true,n} \given y_n, \tilde{\sigma}_n)$ for the highlighted black points. The red points are the true y values.
The blue points in Figure \ref{fig:toy}, lower right panel, show the expectation value and $1\sigma$ uncertainty of the posterior belief for all 1024 $y_n$.
When the true values are inferred in this way, the associated variance $\tilde{\sigma}_n^2$ is the convolution of the true distribution variance $t^2$ and the measurement variance $\sigma_n^2$.
The points are colored by their measurement variance $\sigma_n^2$ which shows that noisier measurements are more influenced by the prior and get pulled closer to the center of the MML distribution, our prior. In contrast, more certain measurements remain closer to their measured values $y_n$.
By inferring the true y value from the measured y value in addition to the prior information of the distribution the y value was drawn from, this we have better estimates of the true y value for each point. Bringing the toy model back to our original problem, using the noise-deconvolved CMD as our prior allows us to get better estimates of the parallax for each star.

@@

We make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the Discussion below.
\begin{description}
\item[stationarity] We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.  In detail, this assumption is wrong, of
  course (for example, distant halo stars are different than local
  disk stars), but the stationarity assumption is fairly weak; as long
  as there is support in the high signal-to-noise stars for the types
  of stars seen at low signal-to-noise, the results will not be
  strongly biased, but this assumption is deep and fundamental to this
  work.
\item[big data] We assume that we have large numbers of stars, large enough that
  an empirical-Bayes or maximum-marginalized likelihood is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case of a stellar type with only one (statistically distinct)
  exemplar in the data set (for example, a white dwarf), this assumption will be maximally wrong.
\item[noise model] We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the [INSERT
    CORRECT TERMINOLOGY FROM GAIA HERE] parallax uncertainty estimates
  are correct. Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative
  (it leads to under-deconvolution).
\item[dust] We treat the \panstarrs-based three-dimensional dust maps (\citealt{green15})
  as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical-Bayes model, we iteratively update each star's
  dust correction at its updated, inferred distance. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for either
  color or absolute magnitude.
\item[mixture of Gaussians] We assume that the \cmd\ can be represented with a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128, 512, and 2048 Gaussians in
  different experiments.
\item[no physics] We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact).
\end{description}

>>>>

In this paper we use the noise deconvolved observed \cmd\ as a prior to infer more precise parallax pdfs from \gaia\ measurements.
Understanding this method is challenging because we are using the data both for the prior and the likelihood in our inference.
To ease understanding, here we will demonstrate this method with a simpler toy model.
Instead of inferring more precise parallax pdfs using the 2D distribution of the \cmd\ as our prior, for this simpler model we will infer more precise pdfs of the value $Y$.
We will do this simpler inference using a prior that is also a 2D distribution: a 1D Gaussian in the y-direction, with a running mean $\mu = mx + b$ and some thickness $t$.
Here the thickness of the toy distribution mimics the true thickness of the \cmd\ due to age, mass and metallicity.
The $1\sigma$ contours as well as (noise free) samples from the toy model distribution are shown in Figure \ref{fig:toy}, upper left panel.
Similarly, noisy samples from this distribution are shown in the upper right panel.
Using a representation of the true distribution (which we learn from the noisy data) as our prior, we can infer some posterior belief over $y_{\true,n}$, a true sample of the distribution, from $y_n$, a noisy sample of the distribution.

Setting up the problem in more detail, we have the true samples from the underlying distribution and the noisy samples

@@

Here is a good point to step back and acknowledge that our method is a bit strange. We are using the set of data to determine the noise-deconvolved distribution, which is an estimate of the true distribution, and then using that noise-deconvolved distribution to make inferences on the data. We are technically using the data twice!To be fully, technically correct, we should learn the true distribution while leaving one datum out, the one datum we want to infer the true y value for\footnote{DWH: Remember what this brain dump is for}. Leaving one datum out to estimate the prior and then inferring the posterior for that datum is way too expensive for the 1024 points in our toy model, and also for the 1.2 million stars in our actual data set. Luckily with large data sets, even at 1024, 1 datum is $\sim0.1\%$ of your dataset and therefore leaving that point out will not change the MML parameters of your noise-deconvolved distribution. With the \gaia\ dataset of over a million stars, we can safely step into the Empirical Bayes regime. (Talk more about emperical bayes here?)

So, returning the inference, we use Bayes theorem to turn the likelihood of the noisy data into a posterior believe over the true y value.
\begin{eqnarray}
p(y_{\true,n} \given y_n, \tilde{\sigma}_n) &=& p(y_n \given y_{\true,n}, \sigma_n)\,p(y_{\true,n}) \\
\mathrm{where}\;p(y_n \given y_{\true,n}, \sigma_n) &=& \mathcal{N}(y_n \given y_{\true,n}, \sigma_n) \\
\mathrm{and}\;p(y_{\true,n}) &=& \mathcal{N}(y_{\true,n} \given \hat{m}\, x_n + \hat{b}, \hat{t})
\label{eq:toyBayes}
\end{eqnarray}
Since everything is Gaussian, this can be done analytically by multiplying Gaussians together. In Figure \ref{fig:toy} lower left panel, the blue points represent the posterior belief of the true value $p(y_{\true,n} \given y_n, \tilde{\sigma}_n)$ for the highlighted black points. The red points are the associated true y values.
In Figure \ref{fig:toy}, lower right panel, the blue points show the expectation value and $1\sigma$ uncertainty of the posterior belief for all 1024 $y_n$.
When the true values are inferred in this way, the associated variance $\tilde{\sigma}_n^2$ is the convolution of the true distribution variance, $t^2$, and the measurement variance, $\sigma_n^2$.
The points are colored by their measurement variance $\sigma_n^2$, which shows that noisier measurements are more influenced by the prior and get pulled closer to the center of the MML distribution, our prior. In contrast, more certain measurements remain closer to their measured values, $y_n$. We now have better estimates of each true y value by inferring each true y value from the measured y value and using an estimate of the true distribution the measured y value was drawn from as the prior. Bringing the toy model back to our original problem, we can get better estimates of the true parallax for each star by inferring the true parallax from the measured parallax and using the noise-deconvolved \cmd, an estimate of the true \cmd, as our prior.

@@

To infer the true parallax to a star using the noise-deconvolved \cmd\ as a prior, we make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the discussion section below.
\begin{description}
\item[stationarity] We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.  In detail, this assumption is wrong, of
  course (for example, distant halo stars are different than local
  disk stars), but the stationarity assumption is fairly weak; as long
  as there is support in the high signal-to-noise stars for the types
  of stars seen at low signal-to-noise, the results will not be
  strongly biased, but this assumption is deep and fundamental to this
  work.
\item[big data] We assume that we have large numbers of stars, large enough that
  an empirical-Bayes or maximum-marginalized likelihood is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case of a stellar type with only one (statistically distinct)
  exemplar in the data set (for example, a white dwarf), this assumption will be maximally wrong.
\item[noise model] We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the [INSERT
    CORRECT TERMINOLOGY FROM GAIA HERE] parallax uncertainty estimates
  are correct. Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative
  (it leads to under-deconvolution).
\item[dust] We treat the \panstarrs-based three-dimensional dust maps (\citealt{green15})
  as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical-Bayes model, we iteratively update each star's
  dust correction at its updated, inferred distance. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for either
  color or absolute magnitude.
\item[mixture of Gaussians] We assume that the \cmd\ can be represented with a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128, 512, and 2048 Gaussians in
  different experiments.
\item[no physics] We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact).
\end{description}
